preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-15 Manuscript/manuscript.Rmd
Methods polishing

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

In Hi-C, each chromosome is binned into non-overlapping regions of length *r*, typically 5 kb and above. The *r* parameter defines the resolution of Hi-C data. Here, we designed a strategy called _shifted binning_ that partitions the genome into regions of the same length *r*, but with middle points corresponding to boundaries defined by the original binning.

@@

Cell line-specific genomic annotations were used to build the predictor space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness present in the distributions of genomic distances. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques including *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the SMOTE algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the predictor space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination (RFE). In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

To investigate whether we could alleviate resolution limitations of conventional domain calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution. To demonstrate, we implemented an optimally trained RF model using the framework described above. That is, a random forest model was built on the optimal combination of predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements) to maximally accurately predict boundaries at base _Hi-C resolution level_.

To precisely identify boundary locations, we first construct a _base-level resolution_ predictor space for a particular chromosome *i*. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary point. Here, $n_{i}$ represents the length of chromosome *i*. A threshold (*t*) is specified on $p_{n_{i}}$, such that base coordinates with $p_{n_{i}} \ge t$ are extracted. Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between bases with $p_{n_{i}} \ge t$. The resulting clusters of highly predictive bases identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (**Clustering Large Applications**) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base coordinate within each clustered PTBR.

The DBSCAN algorithm takes two parameters, *MinPts* and *eps* ($\epsilon$). The *MinPts* parameter was set to the recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. To decide on the optimal value of *t* and $\epsilon$ in *preciseTAD*, we considered the normalized enrichment of flanked predicted boundaries, defined as

@@

Next, we assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the **computeMatrix** and **plotHeatmap** tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon Rank-Sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both Arrowhead and Peakachu boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The alpha level of significance was set to 0.05 for all statistical tests.

>>>>

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

In Hi-C, each chromosome is binned into non-overlapping regions of length *r*, typically, 5 kb and above. The *r* parameter defines the resolution of Hi-C data. Here, we designed a strategy called _shifted binning_ that partitions the genome into regions of the same length *r*, but with middle points corresponding to boundaries defined by the original binning.

@@

Cell-line-specific genomic annotations were used to build the predictor space. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*) (Figure 1B). A ($log_2 + 1$)-transformation of distance was used to account for the skewness of the distance distributions. Models built using a *Peak Signal Strength* predictor space were only composed of histone modifications and transcription factor binding sites because BroadHMM chromatin segmentations lack signal values.

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques. These include *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the SMOTE algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the predictor space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination (RFE). In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features incremented by the power of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

To investigate whether we could alleviate resolution limitations of conventional domain calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution. A random forest model was built on the optimal combination of predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements) to maximally accurately predict boundaries at the _Hi-C resolution level_.

To precisely identify boundary locations, we first constructed a _base-level resolution_ predictor space for a particular chromosome *i*. We then apply our optimally trained RF model on the base-level predictor space to extract the probability vector, $p_{n_{i}}$, denoting each base's probability of being a boundary. Here, $n_{i}$ represents the length of chromosome *i*. A threshold (*t*) is specified on $p_{n_{i}}$, such that base coordinates with $p_{n_{i}} \ge t$ are extracted. Next, density-based spatial clustering of applications with noise (DBSCAN; version 1.1-5) is applied to the matrix of pairwise genomic distances between bases with $p_{n_{i}} \ge t$. The resulting clusters of highly predictive bases identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base coordinate within each clustered PTBR.

The DBSCAN algorithm has two parameters, *MinPts* and *eps* ($\epsilon$). The *MinPts* parameter was set to the recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. To decide on the optimal value of *t* and $\epsilon$ in *preciseTAD*, we considered the normalized enrichment $N_{e}$ of flanked predicted boundaries, defined as

@@

Next, we assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* and *plotHeatmap* tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon Rank-Sum tests were used to compare chromosome-specific Jaccard indices across cell lines, between *preciseTAD* boundaries and both Arrowhead and Peakachu boundaries, respectively. All statistical analyses were performed in R version 4.0.1. The significance level was set to 0.05 for all statistical tests.
